Validated Model Counting Data \\ Version of \today

### 2023-02-27 doc/mc-data.tex
Data on optimization benefits

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \input{data-formatted/defining+total}
    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};

@@

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
                              x post scale=2.0, y post scale=2.2,
                              xmode=log,xmin=0.01,xmax=10000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
    \input{data-formatted/time-d4-verify}
    \node[left] at (axis cs: 0.09,0.12) {$1{}.0\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};

@@

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1e9, 
      legend style={at={(0.40,0.98)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.3,
                              xmode=log,xmin=100,xmax=1e7, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-lemma-nolemma}
    \input{data-formatted/clauses-nosplit-split}
    \legend{
      \scriptsize \textsf{Without lemmas},
      \scriptsize \textsf{Without literal grouping},
    }
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e8,1e8)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};

    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};

@@

>>>>

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \input{data-formatted/defining+total}
    \node[left] at (axis cs: 900,1200) {$1\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};

@@

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
                              x post scale=2.0, y post scale=2.2,
                              xmode=log,xmin=0.01,xmax=10000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
    \input{data-formatted/time-d4-verify}
    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};

@@

\section*{Benefits of Optimizations}

Two optimizations were implemented within the CRAT framework:
\begin{description}
\item[Lemmas:] The justification proof must show that the conjunction
  of the input clauses implies the extension variable associated with
  the root node.  It does so by showing that this property holds for
  every node and for every path from the root that node, relative to
  the partial assignment given by that path.  This could potentially
  require enumerating exponential numbers of paths to some nodes,
  effectively converting the graph into a tree.  Instead, the proof
  can contain a {\em lemma} for each node having indegree greater than
  one, showing that the root node is implied by a set of clauses.
  Some of these are the original input clauses, while some are {\em
    synthetic clauses} representing input clauses that have been
  reduced by eliminating false literals.  Given such a proof, the
  lemma can be instantiated for each parent node.  Proofs with lemmas
  incur some overhead---the synthetic clauses are declared as POG
  product nodes, and additional clauses must be added to show that,
  under a given context, the synthetic clauses are implied by some
  input or containing synthetic clause.  On the other hand, they
  guarantee that no node in the graph is the root of more than one
  justication proof.
\item[Literal grouping:] The generated POG representation will
  typically have a number of product nodes forming the conjunction of
  tens, or even hundreds, of literals.  These literals represent ones
  that we either implied by unit propagation or were added when
  assigning the complement to the corresponding variable led to a
  conflict.  Justifying this product node requires showing that each
  of these literals must hold under the given context.  When this
  cannot be done by unit propagation, the program must invoke a SAT
  solver to prove that the complemented value yields a conflict.
  Rather than doing so for each literal individually, the prover can
  create a new product node consisting of the literals that have not
  yet been validated and then prove that falsifying the extension
  variable associated with the product node yields a conflict.  
\end{description}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
      legend style={at={(0.40,0.98)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.3,
                              xmode=log,xmin=1000,xmax=1e7, 
                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-lemma-nolemma}
    \input{data-formatted/clauses-nosplit-split}
    \input{data-formatted/clauses-optimized-unoptimized}    
    \legend{
      \scriptsize \textsf{Without lemmas},
      \scriptsize \textsf{Without literal grouping},
      \scriptsize \textsf{Without either optimization},
    }
    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};

    \node[left] at (axis cs: 5000,7000) {$1\times$};
    \node[left] at (axis cs: 5000,70000) {$10\times$};
    \node[left] at (axis cs: 5000,700000) {$100\times$};

@@

We conducted a set of experiments using the 80 benchmark instances for
which the total number of proof clauses when both optimizations are
performed was limited to $10^7$.  We then made three runs---with each
of the two optimizations disabled individually and in combination.
Figure \ref{fig:defining:unoptimized} illustrates the effect of the
two optimizations.  Each set of points shows the number of clauses
when both optimizations are performed for the $X$ value and the number
with one or both of these optimizations disabled as the $Y$ value.

The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
\begin{description}
\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
\end{description}
Put another way: None of the 80 benchmarks require more than $10^7$
proof clauses when both optimizations are used.  Without lemmas, 19
rise above that threshold.  Without literal merging, 7 due.  Without
either, 25 rise above that threshold and 2 even exceed $10^8$ proof
clauses.
